---
title: 'CoMM: a collaborative mixed model to dissecting genetic contributions to complex
  traits by leveraging regulatory information'
author: "Jin Liu"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction
This vignette provides an introduction to the `CoMM` package. R package `CoMM` implements CoMM, a collaborative mixed model to dissecting genetic contributions to complex traits by leveraging regulatory information.
The package can be installed with the command:

`library(devtools)`

`install_github("gordonliu810822/CoMM")`

The package can be loaded with the command:
```{r}
library("CoMM")
```

### Fit CoMM using simulated data

We first generate genotype data using function _genRawGeno_:
```{r}
library(mvtnorm)
L = 1; M = 100; rho =0.5
n1 = 350; n2 = 5000;
maf = runif(M,0.05,0.5)
X = genRawGeno(maf, L, M, rho, n1 + n2);
```

Then, effect sizes are generated from standard Gaussian distribution with sparse structure:
```{r}
beta_prop = 0.2;
b = numeric(M);
m = M * beta_prop;
b[sample(M,m)] = rnorm(m);
```
Subsequently, the gene expression `y` is generated by controlling cellular heritability at prespecified level (`h2y`):
```{r}
h2y = 0.05;
b0 = 6;
y0 <- X%*%b + b0;
y  <- y0 + (as.vector(var(y0)*(1-h2y)/h2y))^0.5*rnorm(n1+n2);
```

Finally, the phenotype data is generated as the generative model of CoMM with a prespecified trait heritability (`h2`) as:
```{r}
h2 = 0.001;
y1 <- y[1:n1]
X1 <- X[1:n1,]
y2 <- y0[(n1+1):(n1+n2)]
X2 <- X[(n1+1):(n1+n2),]
alpha0 <- 3  
alpha <- 0.3
sz2 <- var(y2*alpha) * ((1-h2)/h2)
z <- alpha0 + y2*alpha + rnorm(n2,0,sqrt(sz2))
```

The genotype data `X1` and `X2` are normalized as 
```{r}
y = y1;
mean.x1 = apply(X1,2,mean);
x1m = sweep(X1,2,mean.x1);
std.x1 = apply(x1m,2,sd)
x1p = sweep(x1m,2,std.x1,"/");
x1p = x1p/sqrt(dim(x1p)[2])

mean.x2 = apply(X2,2,mean);
x2m = sweep(X2,2,mean.x2);
std.x2 = apply(x2m,2,sd)
x2p = sweep(x2m,2,std.x2,"/");
x2p = x2p/sqrt(dim(x2p)[2])

w2 = matrix(rep(1,n2),ncol=1);
w1 = matrix(rep(1,n1),ncol=1);
```
Initilize the parameters by using linear mixed model (function *lmm_pxem*, LMM implemented using [PX-EM algorithm](https://www.jstor.org/stable/2337481?seq=1#page_scan_tab_contents)):
```{r}
fm0 = lmm_pxem(y, w1,x1p, 100)
sigma2beta =fm0$sigma2beta;
sigma2y =fm0$sigma2y;
beta0 = fm0$beta0;
```
Fit CoMM w/ and w/o constraint that alpha = 0 as
```{r}
fmHa = CoMM_covar_pxem(y, z, x1p, x2p, w1, w2,constr = 0);
fmH0 = CoMM_covar_pxem(y, z, x1p, x2p, w1, w2,constr = 1);
loglikHa = max(fmHa$loglik,na.rm=T)
loglikH0 = max(fmH0$loglik,na.rm=T)
tstat = 2 * (loglikHa - loglikH0);
pval = pchisq(tstat,1,lower.tail=F)
alpha_hat = fmHa$alpha 
```

### Fit CoMM using GWAS and eQTL data

The example of running CoMM using GWAS and eQTL data in plink binary format
```{r}
file1 = "1000G.EUR.QC.1";
file2 = "NFBC_filter_mph10";
file3 = "Geuvadis_gene_expression_qn.txt";
file4 = "";
file5 = "pc5_NFBC_filter_mph10.txt";
whichPheno = 1;
bw = 500000;
```
Here, file1 is the prefix for eQTL genotype data in plink binary format, file2 is the GWAS data in plink binary format, file3 is the gene expression file with extended name, file4 and file5 are covariates file for eQTL and GWAS data, respectively. Then run `fm = CoMM_testing_run(file1,file2,file3, file4,file5, whichPheno, bw);`. For gene expresion file, it must have the following format (rows for genes and columns for individuais):
```{r, echo=FALSE, results='asis'}
knitr::kable(ge[,1:8])
```

To make 'CoMM' further speeding, we implement multiple thread version of 'CoMM' by just run `fm = CoMM_testing_run_mt(file1,file2,file3, file4,file5, whichPheno, bw, coreNum);` where `coreNum = 24` is the number of cores in your CPU.

### Figures

The following data and codes are used to produce one of the figures in the Yang et al. (2018).

```{r, fig.width=7, fig.height=4, echo=FALSE}
dat_rej = dat[[3]];
dat_rej$h2z=paste("",dat_rej$h2,sep="")
dat_rej$Power = dat_rej$rej_prop
dat_rej$Sparsity = dat_rej$beta_prop
dat_rej$sd_rej = as.numeric(as.character(dat_rej$sd_rej))
dat_rej = dat_rej[dat_rej$Method!="2-stage:AUDI",]
library(plyr)
dat_rej$Method=revalue(dat_rej$Method, c("AUDI"="CoMM"))
dat_rej$Method=droplevels(dat_rej$Method)
rho = 0.5; n2 = 8000;
t1e_rej = dat_rej[dat_rej$RhoX==rho&dat_rej$n2==n2,]

t1e_rej$h2z = factor(t1e_rej$h2z)
t1e_rej$h2y = factor(t1e_rej$h2y)
t1e_rej$Sparsity = factor(t1e_rej$Sparsity)
t1e_rej$n2 = factor(t1e_rej$n2)
t1e_rej$Method <- ordered(t1e_rej$Method, levels = c("CoMM","2-stage:Ridge","2-stage:Enet","SKAT"))
t1e_rej$Power = as.numeric(as.character((t1e_rej$Power)))

t1e_rej$h2y2 <- factor(t1e_rej$h2y, labels = c("h[C]^2==0.01", "h[C]^2==0.03", "h[C]^2==0.05", "h[C]^2==0.07", "h[C]^2==0.09"))
t1e_rej$h2z2 <- factor(t1e_rej$h2z, labels = c("h[T]^2==0", "h[T]^2==0.001", "h[T]^2==0.002", "h[T]^2==0.003"))

library(ggplot2)
ggplot(t1e_rej, aes(x = Sparsity, y = Power,fill = Method))+
geom_bar(stat="identity", position=position_dodge())+
geom_errorbar(aes(ymin=Power-sd_rej, ymax=Power+sd_rej), width=.2,
                 position=position_dodge(.9)) +
facet_grid(h2z2~h2y2,labeller = label_parsed,scales = "free_y") + 
geom_hline(yintercept=0.05,colour="orange",linetype="dashed")

```


